Experience Report: Haskell in Computational Biology

### 2012-03-07 mrfy_experience_report.tex
edits to roughly first 3 columns

\documentclass[preprint,nocopyrightspace,times]{sigplanconf}

@@

@@

@@

Computational biologists detect homologies by building 
algorithms which, given an amino-acid sequence,\remark{This sequence
  (the sequence ``under test'') needs a name. ---NR}
find known proteins of
similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}.\remark{citation needed?}
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm, can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

\section{What we did}

\subsection{The Viterbi Algorithm}

A biologist may use software such as HMMER or MRFy to test a hypothesis about 
the function of a single, newly discovered protein, or she may use the software 
to scan an entire newly sequenced genome against a library of known protein 
structures.
This latter case demands particularly high-performance software.

In typical use, HMMER builds its structural template -- its Markov random field 
-- from a group of proteins that share common function and structure.
A 
biologist puts these protein sequences into a correspondence, or alignment, 
typically using structural alignment software.
The columns of the alignment 
(Figure~\ref{alignment}) may contain gaps, indicating that a particular protein 
does not have an amino acid, or residue, participating in that position.
Gaps 
in a few sequences at a position likely correspond to evolutionary deletions, 
while gaps in many sequences at a position may correspond to evolutionary 
insertions.
A position that has residues in a majority of the sequences becomes 
a \textit{match} state in the hidden Markov model; this position may see a 
variety of amino acids due to evolutionary substitution, but it still 
corresponds to a position that a homologous protein should most likely fill.

@@

\caption{Protein sequences are put into alignment based on structural or 
sequence similarities.
Each column of the alignment corresponds to a node of 
the hidden Markov model; columns with many gaps will induce highly probable 
insertion states in the HMM, while columns with fewer gaps will induce 
more-probable deletion and match states.
Columns with strong consensus towards 
a particular residue will induce match states with high emission probability 
for that residue.}\label{alignment} 
\end{figure}

The resulting match states, as well as \textit{insertion} states that allow 
additional residues to appear in a protein sequence, and \textit{deletion} 
states that allow match states to be skipped, form a directed graph, the hidden 
Markov model (Figure~\ref{plan7}).
Each match or insertion state has an 
associated table of emission probabilities, which correspond to the probability 
of seeing each specific amino acid letter at a given position.
Each edge in the 
graph has a weight which is the probability of the associated state transition.

Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
which correspond to columns of the original alignment.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.

@@

\caption{The 
directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
Markov model.
Only 7 of the possible 9 state transitions are allowed, which 
simplifies the dynamic programming.
Insertion states represent mutations that 
insert an amino acid, deletion states represent mutations that lose an amino 
acid, and match states represent either conserved amino acids (those with no 
changes) or mutations that substitute one amino acid for another.
Each match or 
insertion state has an associated table of emission probabilities (here 
illustrated only for the first two states).
Triplets of insertion, deletion, 
and match states are grouped into ``nodes,'' represented here with dashed 
ovals; each node corresponds to a column in the original alignment used to 
train the HMM.}\label{plan7} \end{figure}

If a biologist has trained a hidden Markov model on a group of related 
proteins, she may next use HMMER to determine the likelihood that a new 
protein, whose sequence is known but structure and function are not, shares 
structure and evolutionary history with the proteins that comprise the hidden 
Markov model.
HMMER attempts to compute the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse is a mapping of the letters of a query sequence onto the states 
of the model; equivalently, a parse is a \textit{path} through the edges of the 
graph.
Since each edge has a weight which is a probability, and each emitting 
state (match or insertion) has a table of emission probabilities, the 
probability of a particular path through the model is simply the product of the 
associated probabilities.
This path and its associated probability are computed 
using the Viterbi algorithm~\cite{Viterbi:1967}.

>>>>

\documentclass[preprint,nonatbib,blockstyle,nocopyrightspace,times]{sigplanconf}

@@

\newcommand\figref[1]{Figure~\ref{#1}}

@@

\usepackage[authoryear]{natbib}
\bibpunct();A{},
\let\cite\citep
\let\citeyearnopar=\citeyear
\let\citeyear=\citeyearpar

@@

Computational biologists detect homologies by building 
algorithms which, given an amino-acid sequence,\remark{This sequence
  (the sequence ``under test'') needs a name.  (Urk! I see below it is
  called the ``query sequence.''  This name needs to come in earlier.) ---NR}
find known proteins of
similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}.\remark{citation needed?}
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

\section{The Software}

%% \subsection{The Viterbi Algorithm}

Homology-detection software most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
\remark{Drop this sentence? This latter case demands particularly high-performance software.}
In~both cases, the software is ``trained'' \remark{do we want to talk ``Training''?}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}.
\remark{What does the alignment signify?  Biological significance?
  Computational significance?}
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.
A~column
may contain gaps, which indicate
indicating certain protein have no amino acid participating in that
position.
\remark{The word ``position'' hasn't been defined, and can we get away
  without ``residue''?  (If~we need ``residue,'' let's give it a
  proper introduction.}
If~a column contains just one or two gaps, the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
deleted an amino acid from that position (\figref{alignment}).
If~a column contains \emph{mostly} gaps, 
the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
\emph{inserted} an amino acid in that position (\figref{alignment}).

The protein alignment is used to train a \emph{hidden-Markov model}.
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be structurally related to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
For a given column~$C$ of the alignment, the hidden-Markov model has a
\emph{match state}.
The match state determines the probability that the sequence under
test has a particular amino acid in column~$C$.
The match state assigns high probability to the amino acids that
appear in column~$C$, and a low but nonzero probability to other amino
acids.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$C$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$C$,
and low probability if there are few gaps in column~$C$.
\item
Between every adjacent pair of columns in the alignment,\remark{This isn't right}
the hidden-Markov model has an
\emph{insertion state}.
The insertion state determines the probability that the sequence under
test has an ``extra'' amino acid that does not correspond to any
column in the alignment.
The insertion state assigns high probability to \emph{[some amino
    acids]}, and low probability to \emph{[other amino acids]}.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
\remark{LOST: Each edge in the 
graph has a weight which is the probability of the associated state transition.}
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.
\remark{The evolution seems like a distraction to me here, and we hit
  it above.  Can we simply drop this sentence?}
\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}

@@

\emph{[Not sure we need this text: 
Protein sequences are put into alignment based on structural or 
sequence similarities.
Each column of the alignment corresponds to a node of 
the hidden Markov model; columns with many gaps will induce highly probable 
insertion states in the HMM, while columns with fewer gaps will induce 
more-probable deletion and match states.
Columns with strong consensus towards 
a particular residue will induce match states with high emission probability 
for that residue.]}

\caption{A~structural alignment of four proteins}
\label{alignment} 
\end{figure}

@@

This directed graph represents the states of a
hidden-Markov model in the ``Plan7'' format defined by the HMMER
software package.
%% Only 7 of the possible 9 state transitions are allowed, which 
%% simplifies the dynamic programming.
Each insertion state~($I$) represents a potential mutation that 
inserts an amino acid;
each deletion state~($D$) represents a potental mutation that
deletes an amino acid;
and each match state~($M$) represents either
an unmutated amino acid or a potential mutation that
substitutes one amino acid for another.
A~match state or an insertion state includes a table that assigns a
nonzero probability to each amino acid.
(The figure shows tables for the first two states only.)
A~triplet of an insertion state, deletion state, 
and match state formas a \emph{node}, which corresponds to a column in
the alignment used to train the model.
(Each node is shown in a dashed oval.)

\caption{A hidden-Markov model trained from a four-column alignment}

\label{plan7} \end{figure}

A hidden-Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden-Markov 
model.
A parse maps each amino acid of the sequence under test onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967}.
